---
title: "prep_quest"
author: "Niamh MacSweeney"
date: "25/01/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

#### Introduction

The purpose of this script is to quality control and process the
questionnaire data from the Irritability project:

#### SetUp

The raw (untabulated) questionnaire data is located in:
/Volumes/hwhalley-adol-imaging/Questionnaire_data/2021_07_05_download.xlsx

See /README_ISSF_questionnaire_data.rtf for details on questionnaires
included (this will also aid searching the unified questionnaire data to
extract measures of interest)

Save tabulated (extracted from unified) in:
/Volumes/hwhalley-adol-imaging/Questionnaire_data/Tabulated_data.

This is a good tutorial on regex (handy for tidying
variables):<https://www.jumpingrivers.com/blog/regular-expressions-every-r-programmer-should-know/>

\*\*\*Note: Liana has already extracted the following measures and these
are available in tabulated form: 1. Brief Resilience Scale (BRS) 2. GCOS
(General Causality Orientations Scale ) 3. PCASS (Perceived Choice and
Awareness of Self Scale) 4. PHQ-9 (Depression measure) 4. RCADS (Revised
Children's Anxiety and Depression Scale)

```{r, setup }

library(tidyverse)
library(dplyr)
library(ggplot2)
library(stringr) #for regular expressions
library(gridExtra)
library(RColorBrewer)
library(ggpubr) #for scatter plots corr. coeff


setwd("/Volumes/hwhalley-adol-imaging/irritability_project/prep") #set working directory
getwd()

```

Load in raw questionnaire data.

The behaviour, mood and cognition questionnaire data is located in
post_scan_3.

Note: the data is located in a different wd to this .Rmd file so we will
"hardcode" the data into the script.

```{r, load unified untabulated dataframe}

df <-rio::import("/Volumes/hwhalley-adol-imaging/Questionnaire_data/2021_07_05_download/results-for-post-scan-3-2021-07-05-1537.csv")

PHQ_9 <- rio::import("/Volumes/hwhalley-adol-imaging/Questionnaire_data/Tabulated_data/PHQ9.csv")

RCADS <- rio::import("/Volumes/hwhalley-adol-imaging/Questionnaire_data/Tabulated_data/RCADS.csv")

```

\#\#\#\#Inspect and tidy up variables

Please consult ISSF_data_dictionary to get the column numbers and names
of your variables of interest. Note: Due to the set up on Online Survey,
the introductory question for each measure was given its own column. For
ease of scripting, this column will be removed before giving short names
to each element name. See ARI code chunk below as an example.

e.g.,

To make key column names more intuitive (e.g., subject id) we will
rename the columns.

```{r, inspect variables and tidy}

class(df) #check class, should = data.frame
str(df) #check data structure

#rename subject id column
#it would be easier to do this using column number but to aid reproducibility, we have used the full column names. 

df <- df %>% 
  rename(studyID = "1. Please enter your unique four-digit study code that you will have been given by the research team:")

```

\#\#\#\#\#Affective Reactivity Index

See this Stringaris et al., 2012 paper for
reference:<https://www.kcl.ac.uk/ioppn/depts/cap/research/moodlab/stringaris-ari-paper.pdf>

#### ARI STEP ONE: Scoring and export to tabulated folder

```{r, extract ARI columns and make new df}
df <- df %>% 
  select(- "53. In the last six months and compared to others of the same age, how well does each of the following statements describe your behavior/feelings?") %>% #Remove question column
  rename(ari_1 = "53.1. 1. I am easily annoyed by others", #rename columns
         ari_2 = "53.2. 2. I often lose my temper",
         ari_3 = "53.3. 3. I stay angry for a long time",
         ari_4 = "53.4. 4. I am angry most of the time",
         ari_5 = "53.5. 5. I get angry frequently",
         ari_6 =  "53.6. 6. I lose my temper easily",
         ari_7 = "53.7. 7. Overall, my irritability causes me problems.")

ARI <- df %>% 
  select("studyID","ari_1":"ari_7") #make ARI dataframe

str(ARI) #check data class (=character) so we need to change to numeric

#write function that specifies how we would like to recode character variables to numeric. 
temp_fn <- function(x){
  case_when(x == 'Not true'  ~ 0,
            x == 'Somewhat true' ~ 1,
            x == 'Certainly true'  ~ 2)
}
 ARI <- ARI %>%
  mutate_at(.vars  = vars(c("ari_1", "ari_2", "ari_3", "ari_4", "ari_5", "ari_6", "ari_7")), 
            .funs = temp_fn)

 #tidier code would be better here but get NAs when using c("ari_1": "ari_7") so leaving as is for now. 
 
 #### SCORE ARI ####
 
 #ARI total score = sum of first 6 items (i.e., ari_1 to ari_6)
 #ari_7 is a measure of impairment and is analysed separately. 
 
 ARI$ari_tot <-rowSums(ARI[,c("ari_1", "ari_2", "ari_3", "ari_4", "ari_5", "ari_6")], na.rm=T)
 
 ARI <- ARI %>%  relocate(ari_tot ,.after = studyID) #rearrange cols so ari_tot is after ID
 
 #output tabulated ARI data
 write.csv(ARI, "/Volumes/hwhalley-adol-imaging/Questionnaire_data/Tabulated_data/ARI.csv")
```

#### ARI STEP TWO: Inspect ARI distribution

Range of possible scores = 0 - 12

```{r, extract ARI columns and make new df}
#inspect ARI scores #seem to be low irritability scores over all. 
 
summary(ARI$ari_tot) # get sum stats: mean = 3.226, range = 0-10
sd(ARI$ari_tot) #SD = 2.642

#Let's plot the ARI data
#Looks like most people are on the lower end of irritability

ARI_hist <- ARI %>% 
  ggplot(aes(x = ari_tot)) +
  geom_histogram(bins = 10,
                colour = "skyblue4",
                 fill = "skyblue3") +
  labs(title = "Affective Reactivity Index (ARI)", x = "ARI Total") +
                scale_x_continuous(breaks=seq(0, 12, 2), limits = c(0,12))
print(ARI_hist)

ggsave("/Volumes/hwhalley-adol-imaging/irritability_project/figs/ARI_hist.png")

```

##### ARI Average Total Score

From Stringaris et al., 2012:
<https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&cad=rja&uact=8&ved=2ahUKEwjO9I_Eyqf2AhVJTcAKHZZHBkMQFnoECBwQAQ&url=https%3A%2F%2Fwww.psychiatry.org%2FFile%2520Library%2FPsychiatrists%2FPractice%2FDSM%2FAPA_DSM5_Level-2-Irritability-Child-Age-11-to-17.pdf&usg=AOvVaw0KAwkEkwgxzKY7cDgnrb0A>

Higher scores indicate greater severity of irritability. In addition,
the clinician is asked to calculate and use the average total score. The
average total score reduces the overall score to a 3-point scale, which
allows the clinician to think of the child's irritability in terms of
none (0), mild-moderate (1), or moderate-severe (2). The use of the
average total score was found to be reliable, easy to use, and
clinically useful to the clinicians in the DSM-5 Field Trials.2

The average total score is calculated by dividing the raw total score by
6 (i.e., the first 6 items on the measure).

First, generate average total score:

None = 0 mild-moderate = 1 Moderate severe = 2

Then, use average total score categories to generate binary variable:
Average total score = 0 = not irritated, Group = 0 Average total score
\<= 1 = irritated, mild-moderate or moderate severe irritability. Group
= 1

\*\*\*Note: We may revisit this coding and change it to an ordinal
measure (e.g., range 0-2) rather than binary (e.g., 0-1) but for the
time being, we will use a binary measure to keep it consistent with the
depressio measure.

```{r, ARI average total score}

ARI <- ARI %>% 
  mutate(ari_avg_tot = ari_tot/6) #generate average total score

#round to 0 decimal places to get whole number value
ARI$ari_avg_tot <- round(ARI$ari_avg_tot, digits = 0)

#make grouping variable: 0 = not irritated, 1 = irritated
ARI <- ARI %>%
  mutate(irrit_group = case_when(ari_avg_tot == 0 ~ 0,
                              ari_avg_tot >= 1 ~ 1
                              ))

#make variable factor as it is a grouping variable
ARI$irrit_group = as.factor(ARI$irrit_group)
#check it worked
str(ARI$irrit_group)
print(ARI$irrit_group)

irrt_df <- data.frame(ARI$irrit_group) #convert to dataframe to get group count
count(irrt_df, ARI.irrit_group) #use count function to get group breakdown 
  
#Breakdown: Not Irritated = 20; Irritated = 11

high_irrit <- ARI %>% 
  select(studyID, irrit_group) %>% 
  filter(irrit_group == 1)

low_irrit <- ARI %>% 
  select(studyID, irrit_group) %>% 
  filter(irrit_group == 0)
         

```

##### PHQ-9: Measure of Depression

See this

PHQ-9 Score Depression Severity 0 -- 4 None-minimal 5 -- 9 Mild\
10 -- 14 Moderate 15 -- 19 Moderately Severe\
20 -- 27 Severe

```{r, PHQ-9 total, continuous measure}
summary(PHQ_9$PHQ_total) # get sum stats: mean = 11.87, range = 3-25
sd(PHQ_9$PHQ_total) #SD = 6.607

#Let's create categories based on PHQ-9 scoring

#write a function to classify the categories
cat_fn <- function(x){
  case_when(
    between(x, 0, 4) ~ 'None-minimal',
    between(x, 5, 9) ~ 'Mild',
    between(x, 10, 14) ~ 'Moderate',
    between(x, 15, 19) ~ 'Moderate-severe',
    between(x, 20, 27) ~ 'Severe')

}

#define object x. Here, we want the function to be applied to PHQ_total
x <- PHQ_9$PHQ_total

#apply function
PHQ_9 <- PHQ_9 %>% 
mutate(PHQ_severity = cat_fn(x))

#count number of participants in each category. 
PHQ_table <- PHQ_9 %>% 
    group_by(PHQ_severity) %>% 
    tally()
print(PHQ_table)

png("/Volumes/hwhalley-adol-imaging/irritability_project/figs/PHQ_table.png",
    height = 100*nrow(PHQ_table), width =200*ncol(PHQ_table))
grid.table(PHQ_table)
dev.off()

#Let's plot the PHQ_total (histogram)

PHQ_tot_hist <- PHQ_9 %>% 
  ggplot(aes(x = PHQ_total)) +
  geom_histogram(bins = 15,
                colour = "skyblue4",
                 fill = "skyblue3") +
  labs(title = "PHQ-9", x = "PHQ Total") +
  scale_x_continuous(breaks=seq(0, 27, 3), limits = c(0,27))

print(PHQ_tot_hist)
```

\#\#\#\#PHQ-9 Binary measure Let's generate a group variable that
classifies participant as either "depressed" or "not depressed". PHQ-9
total score \>= 10 = depressed, indicating moderate to severe
depression; Group = 1 PHQ-9 total score \<= 10 = not depression,
indicating none to mild symptoms of depression. Group = 0

```{r, PHQ-9 binary}

#make grouping variable: 0 = not depressed, 1 = depressed 
PHQ_9 <- PHQ_9 %>%
  mutate(dep_group = case_when(PHQ_total < 10 ~ 0,
                              PHQ_total >= 10 ~ 1
                              ))

#Change grouping variable type from numeric to factor
PHQ_9$dep_group = as.factor(PHQ_9$dep_group)
#check it worked
str(PHQ_9$dep_group)
print(PHQ_9$dep_group)


dep_df <- data.frame(PHQ_9$dep_group) #convert to dataframe to get group count
count(dep_df, PHQ_9.dep_group) 

#Group breakdown
#Not depressed = 14; depressed = 17


#Let's plot the PHQ_severity categories (barplot)
# 
# PHQ_sev_bar <- PHQ_9 %>% 
#   ggplot(aes(x = PHQ_severity) +
#     geom_bar())

```

Correlation between PHQ-9 and ARI

```{r, PHQ-9}

#Merge dfs
working_df <- merge(ARI, PHQ_9, by = "studyID")

corr <- cor.test(working_df$PHQ_total,working_df$ari_tot, method = "pearson")
#r = 0.519

#Plot ARI and PHQ relationship 
#define plot
ari_phq_plot <- ggplot(data = working_df,
                       mapping = aes(x = ari_tot, y = PHQ_total))
#draw plot
ari_phq_plot <- ari_phq_plot +
   geom_point(alpha = 0.7, size=2, aes(color = PHQ_severity),
              fill = "transparent") +
  labs(title = "Depression + Irritability Relationship",
       y= "PHQ-9 Total Score",
       x = "Affective Reactivity Index Total Score",
       colour = "PHQ-9 Severity") +
  geom_smooth(method = "lm") +
  stat_cor(method = "pearson", label.x = 1.5, label.y =  25) +
   scale_color_brewer(palette = "Paired") 

print(ari_phq_plot)
ggsave("/Volumes/hwhalley-adol-imaging/irritability_project/figs/ari_phq_plot.png")


```

###RCADS: Mmeasure of depression 

Extract RCADS Major Depression Score

```{r, RCADS-depression}

RCADS_dep <- RCADS %>% 
  select("studyID", "Major depression")

#check variable type
str(RCADS_dep$`Major depression`) #integer 


```
Correlation between RCADS and PHQ-9 as a sensitivity analysis 

```{r RCADS PHQ-9 correlation}

#add RCADS_dep to working directory using merge. 

working_df <- merge(working_df, RCADS_dep, by = "studyID")

working_df <- working_df %>% 
  rename(RCADS_depression = `Major depression`) #rename RCADS depression measure for clarity 

#let's look at the correlation
corr1 <- cor.test(working_df$PHQ_total,working_df$RCADS_depression, method = "pearson")
#r = 0.856 

#Plot PHQ and RCADS_depression relationship 
#define plot
phq_RCADS_plot <- ggplot(data = working_df,
                       mapping = aes(x = RCADS_depression, y = PHQ_total))
#draw plot
phq_RCADS_plot <- phq_RCADS_plot +
   geom_point(alpha = 0.7, size=2, aes(color = PHQ_severity),
              fill = "transparent") +
  labs(title = "PHQ-9 and RCADS Depression Relationship",
       y= "PHQ-9 Total Score",
       x = "RCADS Depression Score",
       colour = "PHQ-9 Severity") +
  geom_smooth(method = "lm") +
  stat_cor(method = "pearson", label.x = 1.5, label.y =  25) +
   scale_color_brewer(palette = "Paired") 

print(phq_RCADS_plot)

```


#RCADS Binary Measure 

```{r, RCADS}

#make grouping variable: 0 = not depressed, 1 = depressed 
RCADS <- PHQ_9 %>%
  mutate(dep_group = case_when(PHQ_total < 10 ~ 0,
                              PHQ_total >= 10 ~ 1
                              ))

#Change grouping variable type from numeric to factor
PHQ_9$dep_group = as.factor(PHQ_9$dep_group)
#check it worked
str(PHQ_9$dep_group)
print(PHQ_9$dep_group)


dep_df <- data.frame(PHQ_9$dep_group) #convert to dataframe to get group count
count(dep_df, PHQ_9.dep_group) 



```


Extract mental health variables for use in HALFpipe group level analysis

```{r, HALFpipe extract covs}

leida_data <- working_df %>% 
  select(studyID, ari_tot, PHQ_total, irrit_group, dep_group)

leida_data <- leida_data %>% 
  filter(studyID != 1016)

setwd("/Volumes/hwhalley-adol-imaging/irritability_project/data")
write.csv(leida_data,"/Volumes/hwhalley-adol-imaging/irritability_project/data/leida_data.csv", row.names =  FALSE)



```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
