---
title: "prep_LEiDA_data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Volumes/hwhalley-adol-imaging/irritability_project/LEIDA_Toolbox')

library(tidyverse)
```

## Introduction

The purpose of this script is to extract timeseries from HALFpipe output for use in LEiDA Toolbox

For details on LEiDA analysis, please see: https://sites.google.com/view/psychomark/home

We will use the Brainnetome Atlas output from HALFpipe. HALFpipe derivatives can be found here: "/Users/niamhmacsweeney/Documents/HALFpipe_irrit_reduced_features/derivatives/halfpipe"

Time series file name structure for each participant: func/task-Rest/sub-sub%d_task-Rest_feature-corrMatrix_atlas-brainnetomeCombinedDseg_timeseries.tsv

We have moved the timeseries files to a new directory using the command line:

For reference, the code is:

find . -type f -iname "*_task-Rest_feature-corrMatrix_atlas-brainnetomeCombinedDseg_timeseries.tsv" -exec cp {} /Volumes/hwhalley-adol-imaging/irritability_project/data/LEiDA_timeseries_irrit_data/ \;

All of the necessary files are in /Volumes/hwhalley-adol-imaging/irritability_project/data/LEiDA_timeseries_irrit_data/

for the LEiDA toolbox to work, we now need to rename files based on their condition, e.g., whether participant is depressed or not depressed.

```{r, renaming files}

#load in participant dataframe

data <-rio::import("/Volumes/hwhalley-adol-imaging/irritability_project/data/LEiDA_data.csv")

#get list of ids for each group
#dep group 0 = not depressed; 1 = depressed 

dep_ids <- data[data$dep_group == 1, ] 
print(dep_ids$studyID)
nondep_ids <- data[data$dep_group == 0, ] 
print(nondep_ids$studyID)

#change file names manually (not ideal but tried to get working in R and was taking too much time. May revisit later!)



```

