---
title: "prep_demo"
author: "Niamh MacSweeney"
date: "31/01/2022"
output: html_document
---
#### Introduction
The purpose of this script is to quality control and process the demographic data from the Irritability project:

#### SetUp

The raw (untabulated) questionnaire data is located in: /Volumes/hwhalley-adol-imaging/Questionnaire_data/2021_07_05_download.xlsx

See /README_ISSF_questionnaire_data.rtf for details on questionnaires included (this will also aid searching the unified questionnaire data to extract measures of interest)

Save tabulated (extracted from unified) in: /Volumes/hwhalley-adol-imaging/Questionnaire_data/Tabulated_data.

This is a good tutorial on regex (handy for tidying variables):https://www.jumpingrivers.com/blog/regular-expressions-every-r-programmer-should-know/

The demographic data is found in the post-scan-4 batter and contains: age, race, height, weight, family history, medication status, mental health treatment information, puberty questionnaire, smoking and alcohol use, data for SMID


```{r, setup}

library(tidyverse)
library(dplyr)
library(ggplot2)
library(stringr) #for regular expressions
library(lubridate)
library(measurements) #for wrangling meausurement data


setwd("/Volumes/hwhalley-adol-imaging/irritability_project/prep") #set working directory
getwd()

```
Load in raw demographic data -  located in post_scan_3.


Note: the data is located in a different wd to this .Rmd file so we will "hardcode" the data into the script. 
```{r, load in data}

demo_vars <-rio::import("/Volumes/hwhalley-adol-imaging/Questionnaire_data/2021_07_05_download/results-for-post-scan-4-2021-07-05-1537.csv")

```

####Inspect and tidy up variables

We will rename all the variable so that they are easier to read. Please consult ISSF data dictionary (demographics) for details on variables, including exact phrasing of questions. 

```{r, tidy variables}

#rename all variables

demo_vars <- demo_vars %>% 
  rename(studyID = "1. Please enter your unique four-digit study code that you will have been given by the research team:", 
         date_of_birth = "2. Your date of birth", 
         gender = "3. Which gender do you identify with?",
         sex = "4. What is your biological sex?",
         race = "5. How would you describe your ethnic origin?",
         height = "6. What is your current height? Please include the unit of measurement (e.g., centimetres or feet/inches). If you are unsure, please give your best estimate.",
         weight = "7. What is your current weight? Please include the unit of measurement (e.g., kilograms or stones/pounds). If you are unsure, please give your best estimate.",
         postcode = "8. What is your home postcode?",
         family_hist = "9. Does a close family member (parent/sibling/grandparent) have a history of a mental health problem?",
         family_hist_info = "9.a. Please give details of the issue or diagnosis",
         chronic_condition = "10. Do you suffer from a chronic health condition (e.g. chronic respiratory or inflammatory conditions?",
         chronic_condition_info = "10.a. Please give details of the issue or diagnosis",
         med_status = "11. Are you currently taking any medications?",
         med_status_info = "11.a. Please give details on the type of medication, dates and duration of treatment.",
         psych_med_status = "12. Have you taken medication for behavioural or emotional difficulties in the past?",
         psych_med_info = "12.a. Please give details on the type of medication, dates and duration of treatment.",
         psych_ther_status = "13. Are you currently receiving, or have you received psychotherapy counselling or seen a guidance counsellor at school for emotional difficulties?",
         psych_ther_info = "13.a. Please give details on the type, dates and duration of treatment.",
         school_func = "14. Have you needed to take time off school because of low mood?",
         school_func_info = "14.a. For how long?",
         smoking_status = "15. Do you smoke or have you smoked in the past?",
         ecig_status = "16. Do you use e-cigarettes/vapes/vape-pens or have you used them in the past?",
         alcohol_status = "17. Have you ever had an alcoholic drink?",
         alcohol_units = "17.a. During the past week, please record how many units of alcohol you have had. (1 Unit = approximately 1⁄2 a pint of beer, 1 measure of spirit, or 1 glass of wine)",
         alcohol_units_comp = "17.a.i. How does this compare to what you usually drink in a week?",
         puberty_1f = "18. FEMALES: How old were you when you had your period for the first time?",
         puberty_2f = "19. FEMALES: How old were you when you first started to wear a bra?",
         puberty_1m = "20. MALES: How old were you when your voice started to break?",
         
         puberty_2m = "21. MALES: How old were you when you first started to shave your facial hair?",
         
         gp_name = "22. GP Name:",
         
         gp_address = "23. GP Address:",
         
         completion_date = "CompletionDate"
         
         ) 

colnames(demo_vars) #check it worked. 


class(demo_vars) #check class, should = data.frame
str(demo_vars) #check data structure

```


```{r, remove }

```


#Generate age variable from date of birth and completion date
```{r, calculate age}

#DOB and completion date are both character variables
demo_vars %>% 
  select(date_of_birth, completion_date) %>% 
  str()

#First, convert these columns to date format using lubridate function
#They will both be converted to year, month, day and saved as an "interval" class, as per lubridate function. 


demo_vars<- demo_vars %>% 
  mutate(date_of_birth = dmy(date_of_birth),
            completion_date = dmy(completion_date))

demo_vars %>% 
  select(date_of_birth, completion_date) %>% 
  str() #check it worked - it did - both now in Date format!

#calcuate age in years using interval function

#create objects to use with function
dob <- demo_vars$date_of_birth
test_date <- demo_vars$completion_date

#run function
demo_vars$age <- interval(dob, test_date) %>%
  as.numeric('years') #this will give the year as a decimal place
  round(demo_vars$age, digits = 2) #let's round to two decimal places. 

```
#Generate standard weight and height variables 

Height will be in units: cm or ft an/or inches - need to convert all to meters
Weight will be in units: kg or stone and/or pounds - need to convert all to kg. 

BMI calculation requires height and weight in meters and kg. 
####Height 
```{r, tidy height}

print(demo_vars$height)

#use stringr function to remove units

height <- demo_vars$height

#deal with cm units first
demo_vars$height <- str_replace(height, pattern = "cm", replacement = "") #remove cm units (imp. for later conversion to numeric from character vector)

#prepare feet and inches for conversion

#split height column into two new cols called "ft" and "inches" by pattern "ft"
#Note: Where height is given in cm, it will be NA (only worked for extract function, not separate)
demo_vars <- demo_vars %>% 
  extract(col = height, into = c("ft", "inches"), regex = "([[:alnum:]]+)ft([[:alnum:]]+)", 
convert = TRUE, remove = FALSE) %>% 
  mutate(cm = (12*ft + inches)*2.54)


#make ft/inches row in height column = NA to allow later merge. 
demo_vars <- demo_vars %>% 
  mutate(height = case_when(
    str_detect(height, "ft")~ "NA",
    TRUE ~ .$height
    ))

#change height column to numeric 
#Note: you will get warming that NAs introduced by coercion —— this is okay because we converted ft/inches to NAs already in previous step. 
demo_vars$height <- as.numeric(as.character(demo_vars$height))

#check variable classes to ensure numeric before merge
demo_vars %>% 
  select(height, cm) %>% 
str() #both numeric! 

#merge two columns using coalesce function to get final height column 
    demo_vars <- demo_vars %>% mutate(height_tidy = coalesce(height, cm))

#now merge new "cm" column with original height column
demo_vars <- demo_vars %>% 
      mutate(height_tidy = coalesce(height,cm)) #merge
      
demo_vars <- demo_vars %>% 
  mutate(height_m = height_tidy/100) #convert to metres

demo_vars$height_m <- round(demo_vars$height_m, digits = 2) #round to two decimal places
  
 colnames(demo_vars)   #print column names
#remove columns that are no longer needed
    demo_vars <- demo_vars %>% 
      select(-c(height_tidy,height, ft, inches, cm))

colnames(demo_vars) #check it worked
```
 #####Weight 
```{r, weight}
print(demo_vars$height)

#use stringr function to remove units

height <- demo_vars$height

#deal with cm units first
demo_vars$height <- str_replace(height, pattern = "cm", replacement = "") #remove cm units (imp. for later conversion to numeric from character vector)

#prepare feet and inches for conversion

#split height column into two new cols called "ft" and "inches" by pattern "ft"
#Note: Where height is given in cm, it will be NA (only worked for extract function, not separate)
demo_vars <- demo_vars %>% 
  extract(col = height, into = c("ft", "inches"), regex = "([[:alnum:]]+)ft([[:alnum:]]+)", 
convert = TRUE, remove = FALSE) %>% 
  mutate(cm = (12*ft + inches)*2.54)


#make ft/inches row in height column = NA to allow later merge. 
demo_vars <- demo_vars %>% 
  mutate(height = case_when(
    str_detect(height, "ft")~ "NA",
    TRUE ~ .$height
    ))

#change height column to numeric 
#Note: you will get warming that NAs introduced by coercion —— this is okay because we converted ft/inches to NAs already in previous step. 
demo_vars$height <- as.numeric(as.character(demo_vars$height))

#check variable classes to ensure numeric before merge
demo_vars %>% 
  select(height, cm) %>% 
str() #both numeric! 

#merge two columns using coalesce function to get final height column 
    demo_vars <- demo_vars %>% mutate(height_tidy = coalesce(height, cm))

#now merge new "cm" column with original height column
demo_vars <- demo_vars %>% 
      mutate(height_tidy = coalesce(height,cm)) 

 colnames(demo_vars)   #print column names
#remove colulmns that are no longer needed
    demo_vars <- demo_vars %>% 
      select(-c(height, ft, inches, cm))

colnames(demo_vars) #check it worked


```
 
Inspect age variables
```{r, age}


summary(demo_vars$age) # get sum stats: mean = 18.86, range = 17-20
sd(demo_vars$age) #SD=0.83
demo_vars$age <- round(demo_vars$age, digits = 2) #round to two decimal places

```

Inspect and tidy sex variable for group analysis 

```{r, sex}

str(demo_vars$sex)

#Recode sex variable as 0 and 1 so that it works in HALFpipe. 0 = Male, 1 = Female
demo_vars <- demo_vars%>%
  mutate(sex = case_when(sex == "Male" ~ 0,
                              sex == "Female" ~ 1
                              ))

demo_vars$sex = as.factor(demo_vars$sex)
#check it worked
str(demo_vars$sex)
print(demo_vars$sex)


#Get percentage breakdown of sex: Male = 22.6%; Female= 77.4%

demo_vars %>% 
    group_by(sex) %>% 
    summarise(percent = 100 * n() / nrow(demo_vars))

```

### Inspect medication variables

Need to create new binary variable for psychotropic medication history 

0 = medication naive
1 = history of meds or currently taking meds 

```{r, medication vars}

#extract vars of interest from master files 

#double check med_status and med_status_info variables to check for any mismatches. 

demo_vars <- demo_vars %>% 
mutate(med_group = case_when(psych_med_status == "No" ~ 0,
                              psych_med_status == "Yes" ~ 1,
                              )) #recode psych_med_status variable to numeric

```

### Inspect motion variables from HALFpipe output

```{r, motion}

#load data

motion_vars <-rio::import("/Volumes/hwhalley-adol-imaging/irritability")

```



Need to extract covariates of interest for use in further analyses

```{r, generate covariate df for HALFpipe group analysis (dummy run)}

covs <- demo_vars %>% 
  select(studyID, sex, age, med_group)


setwd("/Volumes/hwhalley-adol-imaging/irritability_project/data")
saveRDS(covs,"demo_covs_LEiDA.rds")
```
